Exploring the ages of rotating stars using galactic dynamics: a novel approach to calibrating gyrochronology

### 2020-03-12 paper/conclusion.tex
small typos

In this paper, we used the \vb\ velocity dispersions of stars in the \mct\
catalog to explore the evolution of stellar rotation period as a function of
effective temperature and age.
Our conclusions are as follows:
\begin{itemize}
\item{{\bf Magnetic braking efficiency doesn't always increase with decreasing
mass for K dwarfs.}
% We found that the \citet{angus2019} relation, which is based on the
% period--color relation of the 650 Myr Praesepe cluster, does not correctly
% describe the period--age--\teff\ relation for old stars.
Although at young ages, rotation period is anti-correlated with \teff\ (as
seen in many young open clusters, include Praesepe), at intermediate ages the
relation flattens out and K dwarfs of different masses rotate at the same
rate.
At old ages, cooler K dwarfs spin more rapidly than hotter K dwarfs of the
same age.}

@@

\item{{\bf \sigmavb\ can be used as an age proxy for \kepler\ stars.} In the
    appendix of this paper we demonstrate that the dispersion of velocities in
        the direction of Galactic latitude, \vb, can be used as an age proxy,
        by showing that there is no strong evidence for mass-dependent heating
        in low-mass \kepler\ dwarfs: the velocity dispersions of K and M
        dwarfs, whose main-sequence lifetimes are longer than around 11 Gyrs,
        do not appear to increase with decreasing mass.
Although {\it vertical} velocity, \vz, is a quantity that has been
demonstrated to trace time-dependent orbital heating in the disc of the
Galaxy, most stars with measured rotation periods do not yet have radial
velocities, so we used velocity in the direction of Galactic latitude, \vb\,
as a proxy for \vz.
        Using stars in the GUMS simulation \citep{robin2012}, we showed that
        using \vb\ as a proxy for \vz\ introduces an additional
        pseudo-velocity dispersion, which increases with increasing Galactic
        latitude.
% For this reason we did not attempt to convert \vb\ dispersions into ages using
% an age-velocity dispersion relation.
However, after removing high-latitude ($b>15^\circ$) stars from the sample, we
confirmed that using \vb\ instead of \vz\ does not introduce any
mass-dependent velocity dispersion bias into the sample.
We therefore confirmed that \vb\ velocity dispersion can be used to accurately
rank stars by age.}
% , \ie\ a group of stars with a large velocity dispersion is,
% on average, older than a group of stars with a small velocity dispersion.}

>>>>

In this paper, we used the \vb\ velocity dispersions of stars in the \mct\
catalog to explore the evolution of stellar rotation period as a function of
effective temperature and age.
Our conclusions are as follows:
\begin{itemize}
\item{{\bf Magnetic braking efficiency doesn't always increase with decreasing
mass for K dwarfs.}
% We found that the \citet{angus2019} relation, which is based on the
% period--color relation of the 650 Myr Praesepe cluster, does not correctly
% describe the period--age--\teff\ relation for old stars.
Although at young ages, rotation period is anti-correlated with \teff\ (as
seen in many young open clusters, including Praesepe), at intermediate ages the
relation flattens out and K dwarfs of different masses rotate at the same
rate.
At old ages, cooler K dwarfs spin more rapidly than hotter K dwarfs of the
same age.}

@@

\item{{\bf \sigmavb\ can be used as an age proxy for \kepler\ stars.} In the
    appendix of this paper we demonstrate that the dispersion of velocities in
        the direction of Galactic latitude, \vb, can be used as an age proxy,
        by showing that there is no strong evidence for mass-dependent heating
        in low-mass \kepler\ dwarfs: the velocity dispersions of K and M
        dwarfs, whose main-sequence lifetimes are longer than around 11 Gyrs,
        do not appear to increase with decreasing mass.
Although {\it vertical} velocity, \vz, is a quantity that has been
demonstrated to trace time-dependent orbital heating in the disc of the
Galaxy, most stars with measured rotation periods do not yet have radial
velocities, so we used velocity in the direction of Galactic latitude, \vb\,
as a proxy for \vz.
        Using stars in the GUMS simulation \citep{robin2012}, we showed that
        using \vb\ as a proxy for \vz\ introduces an additional
        velocity scatter, which increases with increasing Galactic
        latitude.
% For this reason we did not attempt to convert \vb\ dispersions into ages using
% an age-velocity dispersion relation.
However, after removing high-latitude ($b>15^\circ$) stars from the sample, we
confirmed that using \vb\ instead of \vz\ does not introduce any
mass-dependent velocity dispersion bias into the sample.
We therefore confirmed that \vb\ velocity dispersion can be used to accurately
rank stars by age.}
% , \ie\ a group of stars with a large velocity dispersion is,
% on average, older than a group of stars with a small velocity dispersion.}

### 2020-03-12 paper/method.tex
small typos

The {\tt Pyia} \citep{price-whelan_2018} and {\tt astropy} \citep{astropy2013,
astropy2018} {\it Python} packages were used to calculate velocities for the
\mct\ sample.
{\tt Pyia} calculates velocity samples from the full \gaia\ uncertainty
covariance matrix via Monte Carlo sampling, thereby accounting for the
covariances between \gaia\ positions, parallaxes and proper motions.
Stars with negative parallaxes, parallax signal-to-noise ratios less than 10,
stars fainter than 16th magnitude, stars with absolute \vb\ uncertainties
greater than 1 \kms\, and stars with galactic latitudes greater than
15\degrees\ (justification provided below) were removed from the sample.
Finally, we removed stars with rotation periods shorter than the main
population of periods, since this area of the period-\teff\ diagram is
sparsely populated.
We removed these rapid rotators by cutting out stars with gyrochronal ages
less than 0.5 Gyr \citep[based on the][gyro-model]{angus2019}, because a 0.5
Gyr gyrochrone\footnote{A gyrochrone is a gyrochronological isochrone, or a
line of constant age in period-\teff, or period-color space.} traces the
bottom edge of the main population of rotation periods.
After these cuts, around 13,000 stars out of the original $\sim$34,000 were
included in the sample.

>>>>

The {\tt Pyia} \citep{price-whelan_2018} and {\tt astropy} \citep{astropy2013,
astropy2018} {\it Python} packages were used to calculate velocities for the
\mct\ sample.
{\tt Pyia} calculates velocity samples from the full \gaia\ uncertainty
covariance matrix via Monte Carlo sampling, thereby accounting for the
covariances between \gaia\ positions, parallaxes and proper motions.
Stars with negative parallaxes, parallax signal-to-noise ratios less than 10,
stars fainter than 16th magnitude, stars with absolute \vb\ uncertainties
greater than 1 \kms\, and stars with galactic latitudes greater than
15\degrees\ (justification provided in the appendix) were removed from the
sample.
Finally, we removed stars with rotation periods shorter than the main
population of periods, since this area of the period-\teff\ diagram is
sparsely populated.
We removed these rapid rotators by cutting out stars with gyrochronal ages
less than 0.5 Gyr \citep[based on the][gyro-model]{angus2019}, because a 0.5
Gyr gyrochrone\footnote{A gyrochrone is a gyrochronological isochrone, or a
line of constant age in period-\teff, or period-color space.} traces the
bottom edge of the main population of rotation periods.
After these cuts, around 13,000 stars out of the original $\sim$34,000 were
included in the sample.
